Gravitational waves from colliding black holes

### 2017-04-28 Report/Draft/draft.tex
Draft
Latest

\subsection{Precession and its astrophysical importance}
Considerable research has been done on studying non-spinning binaries and on non-precessing binaries where the spins are aligned or anti-aligned with the orbital angular momentum $\vec{\hat{L}}$\cite{pe3}, but it is only recently that a more complete study of the parameter space has begun\cite{sloos}\cite{pe_latest}. This is partly down to computational resources, as ignoring spin effects leads to a reduced parameter space and less computationally intensive waveforms. There are a unique set of challenges when considering binaries where the spins of the component black holes are not aligned $\vec{L}$, as due to relativistic effects, these binaries precess around the axis of total angular momentum $\vec{\hat{J}}$, giving a time dependence to the orbital plane of the binary\cite{precess1}\cite{precess2}.
	\includegraphics[scale=0.75]{fig6.jpg}
	\caption{Illustration of a precessing binary system https://phys.org/news/2015-03-insights-black-hole-collisions.html}
This causes the signal in the detector to have an overall amplitude modulation as a function of time due to the change in orientation of the source, and therefore the change in direction of peak emission of GWs. This can be seen in Fig. 7, where we compare a very slightly precessing system with a maximally precessing one. Due to the computational intensity of dealing with both the generation of waveforms and inference process involving precessing systems, significant efforts have been made to reduce the size of the parameter space using the degenerecies between specific spin combinations to parametrise the spins of a binary. The most successful of these is the adoption of two spin parameters\cite{imr}\cite{chip} that describe the whole binary system, where we replace the six spin parameters with two:
\begin{equation}
\chi_{\text{eff}}=\bigg(\frac{\mathbf{S}_1}{m_1}+\frac{\mathbf{S}_2}{m_2}\bigg)\cdot\frac{\mathbf{\hat{L}}}{M}
\end{equation}
and
\begin{equation}
\chi_\text{p}=\frac{1}{B_1m^2_1}\text{max}(B_1S_{1\perp},B_2S_{2\perp})
\end{equation}
where $B1=2+3/(2q)$ and $B2=2+(3q)/2$. These parameters effectively quantify the amount of in-plane and out of plane spin of the total binary, removing large degenerate portions of the parameter space. As different spin configuations within these parameters are effectively degenerate, no loss of information occurs in this re-parametrisation. The recent adoption of this parametrisation in the generation of template waveform banks (the IMRPhenomPv2 waveforms which, we use in this work) has made exploring the precessing parameter space computationally viable, as it reduces the computation time by an order of magnitude. The results of these waveforms agree well when compared with waveforms generated using the full spin parameter space (known as the Spinning Effective One Body Numerical Relativity, or SEOBNR waveforms)\cite{spin}\cite{eob}, so the recent availability of a computationally efficient precessing waveform bank makes this an opportune time to conduct this research. Indeed the first full survey of an isotropic distribution of spins for a large number of simulations (200) has only very recently been published a matter of weeks ago\cite{pe_latest}.
	\includegraphics[scale=1]{fig7.png}
	\caption{Waveforms for three BBH mergers with $m1=30$, $m2=10$ and with spins only in the $x$ direction on the heavier black hole. The inclination is such that the binary is viewed edge-on. LEGEND ON THIS FIGURE IS WRONG, FIX BEFORE SUBMIT}

@@

\subsection{Matching between waveforms}
An extremely useful and computationally cheap way to identify some of these degeneracies is through the process of matching as defined in (13). In the \textit{PyCBC} framework, the match is maximised over extrinsic parameters such as distance and source location that only affect the amplitude of a signal, and so is a useful tool for examining the similarities in the morphologies of different waveforms. A match takes values in the range $[0,1]$, with identical waveforms having a match of 1.
\subsection{Scanning the parameter space}
2D match plots

@@

Describe software injections

@@

Describe the structure of the inference pipeline

>>>>

\subsection{Precession and its astrophysical importance}
Considerable research has been done on studying non-spinning binaries and on non-precessing binaries where the spins are aligned or anti-aligned with the orbital angular momentum $\vec{\hat{L}}$\cite{pe3}, but it is only recently that a more complete study of the parameter space has begun\cite{sloos}\cite{pe_latest}. This is partly down to computational resources, as ignoring spin effects leads to a reduced parameter space and less computationally intensive waveforms. There are a unique set of challenges when considering binaries where the spins of the component black holes are not aligned $\vec{L}$, as due to relativistic effects, these binaries precess around the axis of total angular momentum $\vec{\hat{J}}$, giving a time dependence to the orbital plane of the binary\cite{precess1}\cite{precess2}.
\begin{figure}[h]
	\includegraphics[scale=0.75]{fig6.jpg}
	\centering
	\caption{Illustration of a precessing binary system https://phys.org/news/2015-03-insights-black-hole-collisions.html}
	\centering
\end{figure}
This causes the signal in the detector to have an overall amplitude modulation as a function of time due to the change in orientation of the source, and therefore the change in direction of peak emission of GWs. This can be seen in Fig. 7, where we compare a very slightly precessing system with a maximally precessing one.
\begin{figure}[h]
	\includegraphics[scale=1]{fig7.png}
	\centering
	\caption{Waveforms for three BBH mergers with $m1=30$, $m2=10$ and with spins only in the $x$ direction on the heavier black hole. The inclination is such that the binary is viewed edge-on. LEGEND ON THIS FIGURE IS WRONG, FIX BEFORE SUBMIT}
	\centering
\end{figure}
Due to the computational intensity of dealing with both the generation of waveforms and inference process involving precessing systems, significant efforts have been made to reduce the size of the parameter space using the degenerecies between specific spin combinations to parametrise the spins of a binary. The most successful of these is the adoption of two spin parameters\cite{imr}\cite{chip} that describe the whole binary system, where we replace the six spin parameters with two:
\begin{equation}
\chi_{\text{eff}}=\bigg(\frac{\mathbf{S}_1}{m_1}+\frac{\mathbf{S}_2}{m_2}\bigg)\cdot\frac{\mathbf{\hat{L}}}{M}
\end{equation}
and
\begin{equation}
\chi_\text{p}=\frac{1}{B_1m^2_1}\text{max}(B_1S_{1\perp},B_2S_{2\perp})
\end{equation}
where $B1=2+3/(2q)$ and $B2=2+(3q)/2$. These parameters effectively quantify the amount of in-plane and out of plane spin of the total binary, removing large degenerate portions of the parameter space. As different spin configuations within these parameters are effectively degenerate, no loss of information occurs in this re-parametrisation. The recent adoption of this parametrisation in the generation of template waveform banks (the IMRPhenomPv2 waveforms which, we use in this work) has made exploring the precessing parameter space computationally viable, as it reduces the computation time by an order of magnitude. The results of these waveforms agree well when compared with waveforms generated using the full spin parameter space (known as the Spinning Effective One Body Numerical Relativity, or SEOBNR waveforms)\cite{spin}\cite{eob}, so the recent availability of a computationally efficient precessing waveform bank makes this an opportune time to conduct this research. Indeed the first full survey of an isotropic distribution of spins for a large number of simulations (200) has only very recently been published a matter of weeks ago\cite{pe_latest}.

@@

\subsection{Matching and exploring the parameter space}
An extremely useful and computationally cheap way to identify some of these degeneracies is through the process of matching as defined in (13). In the \textit{PyCBC} framework, the match is maximised over extrinsic parameters such as distance and source location that only affect the amplitude of a signal, and so is a useful tool for examining the similarities in the morphologies of different waveforms. A match takes values in the range $[0,1]$, with identical waveforms having a match of 1. It is also important to note that the inclination parameter has a similar nature of that of the initial phase, in that in matches of non-precessing binaries it can be maximised over as it does not affect signal morphology, however once precession is added, waveforms at different inclinations no longer match, further adding to the complexity of the parameter space. In this section we intend to get some idea of the areas in which we expect the detector to be most sensitive to precession, and those in which precessing signals are intrinsically degenerate with non-precessing waveforms. This is done by incrementally changing waveform parameters and seeing how the match between the waveforms changes as a result. 
	\includegraphics[scale=0.58]{fig9.png}
	\caption{Affect of phase on non-precessing and precessing waveforms. Precession adds an extra degree of freedom in that the morphology of precessing waveforms is different for different phases.}
Fig. 10 shows how the matches between precessing and non-precessing waveforms change with inclination for four different mass ratios. The dark regions of the match is where the signals have significant mismatch, and the waveforms should be well distinguished, and hte light regions are where the waveforms are effectively degenerate. It is apparent that the degeneracy is far stronger for close to equal mass binaries, where only maximally precessing systems at edge-on inclinations will have noticeably different waveforms. Here we plot the 0.97 match contour as it is generally expected that the advanced LIGO detectors should be able to distinguish between waveforms with a match of 0.97. For al but the most extreme mass ratios, a combination of significant in-plane spin (>0.6) and/or strong inclination is necessary to have any significant deviation from a precessing waveform. It is important to note here that as we change the inclination, the overall strength of the signal will drop (with strongest at $\iota=0$), and the matching process does not account for this change in signal strength.
	\includegraphics[scale=0.55]{fig10.png}
	\caption{Matches between precessing and non-precessing waveforms for a range of inclinations}

@@

In this section we describe the process of inserting signals into simulated detector noise, and then perform a full parameter estimation.

@@

A large number of software injections were going to be necessary to generate any interesting results due to the large size and complexity of the parameter space as revealed by the match results. So it was key that they could be performed efficiently and in an organised way. This required the construction of an inference pipeline which formed the bulk of the computational work of the project. This was ultimately one bash script that would create a new folder for each run, generate the injection file and run the inference MCMC.  It also saves the injected parameters as a python dictionary, plots all posteriors overlaid with injected parameters, plots the injected waveform, and the recovered (maximum posterior, or MAP waveform) and injected waveforms overlaid on the whitened detector strain data, and return matches between the MAP and injected waveforms. This meant that all the relevant analysis and data processing needed for each inference run was fully automated. A flow chart of this process is shown in Fig. 12.
This project work was also split between a number of different LIGO clusters, as well as work on several different local PCs, and so the pipeline along with all the relevant match and precession scripts were maintained as part of a GitHub repository which meant that the codes could be developed effectively and without conflict.
	\includegraphics[scale=0.55]{fig12.png}
	\caption{Flow chart of the inference pipeline. The ellipses represent data files that were generated during the pipeline and stored in case further analysis was necessary. The boxes at the bottom represent the final output of the pipeline.}
